---
title: "Compare RFSI method vs RF (ignoring locations) method and evaluate performance"
author: "Narin Polat Öztürk"
format: 
  typst:
    fig-format: png
editor: visual
---

**Task:** Use the meuse data set and focus on “zinc” variable; compare performance of using RFSI method vs simple RF using 5–fold cross-validation: “Is there a difference in accuracy between the two methods?” try adding new covariate layers such as the inundated water and similar and see if this changes results.

## 1. Load the required packages

First, install the packages if they are not already installed.

The packeages are used for the following purposes: 

- sp: provides the Meuse data set.
- meteo: creates nearest-neighbour variables required for RFSI.
- ranger: fits Random Forest models.
- ggplot2: creates figures.
- dplyr and tidyr: data manipulation.
- terra: handling additional raster covariates.

```{r}
#| warning: false
#| message: false
# Libraries
library(sp)
library(meteo)
library(ranger)
library(ggplot2)
library(dplyr)
library(tidyr)
library(terra)
```
# 2. Load the Meuse data

The Meuse data set contains heavy-metal concentrations measured in the floodplain of the Meuse River.

```{r}
# Downloading muse dataset 
data(meuse, package = "sp")
data(meuse.grid, package = "sp")

```

## 3. Define the baseline covariates

The same environmental covariates should be supplied to both RF and RFSI so that the comparison is fair.

```{r}
# Defining the covars
covars <- c("dist", "ffreq", "soil")
names(meuse.grid)

meuse$ffreq <- factor(meuse$ffreq)
meuse$soil <- factor(meuse$soil)

meuse.grid$ffreq <- factor(meuse.grid$ffreq, levels = levels(meuse$ffreq))
meuse.grid$soil <- factor(meuse.grid$soil, levels = levels(meuse$soil))
```
The baseline Random Forest therefore represents a model that uses environmental covariates but does not explicitly use geographical coordinates or neighbouring zinc observations.

## 4. Explore the distribution of zinc

Before fitting the models, inspect the response variable.

```{r}
# Distribution of zinc
summary(meuse$zinc)
ggplot(meuse, aes(x = zinc)) +
  geom_histogram(bins = 25) +
  theme_minimal() +
  labs(x = "Zinc (mg/kg)", 
       y = "Frequency", 
       title = "Distribution of zinc concentrations")
```
Zinc concentrations are expected to show a right-skewed distribution.
For the initial comparison, the models can be fitted directly to the original zinc values. A log-transformed version can later be tested as a sensitivity analysis

## 5. Create 5-fold cross-validation

The observations are divided into five approximately equal groups.

```{r}
# 5-fold cross validation
set.seed(42)

n <- nrow(meuse)

fold_id <- sample(rep(1:5, length.out = n))
table(fold_id)
```
The same folds must be used for both RF and RFSI.

This is important because otherwise differences in model performance could partly result from different train-validation partitions rather than from the modelling method itself.

## 6. Fit the baseline Random Forest

```{r}
# Random Forest (RF)
train_id <- which(fold_id != 1)
test_id  <- which(fold_id == 1)
# training
train <- meuse[train_id, ]
# validation
test <- meuse[test_id, ]
# RF model
rf_model <- ranger(
  zinc ~ dist + ffreq + soil,
  data = train,
  num.trees = 500,
  seed = 42
)

# prediction
rf_pred <- predict(rf_model, data = test)$predictions

head(rf_pred)
```
It also does not use zinc concentrations measured at neighbouring sampling locations.
Therefore, this represents the RF ignoring locations method required in the assignment.

## 7. Construct the spatial predictors for RFSI

The main difference between standard RF and RFSI is that RFSI introduces information from neighbouring observations.

```{r}
# Random Forest Spatial Interpolation (RFS)
nearest_train <- near.obs(
  locations = train,
  locations.x.y = c("x", "y"),
  observations = train,
  observations.x.y = c("x", "y"),
  obs.col = "zinc",
  n.obs = 10,
  rm.dupl = TRUE
)
rm.dupl = TRUE
head(nearest_train)
```

## 8. Calculate spatial predictors for the validation data 

This is the most important part of the cross-validation procedure. For validation locations, neighbouring observations must be searched only within the training data.
```{r}
# Calculating neasrest validation points 
observations = meuse
nearest_test <- near.obs(
  locations = test,
  locations.x.y = c("x", "y"),
  observations = train,
  observations.x.y = c("x", "y"),
  obs.col = "zinc",
  n.obs = 10,
  rm.dupl = TRUE
)
observations = train
```
This prevents validation zinc concentrations from being used as predictors.

If all Meuse observations were used before cross-validation, the validation observations could become neighbours of one another and information would leak from the validation set into the model.

That would artificially improve the apparent performance of RFSI.

## 9. Combine the environmental and spatial predictors 

```{r}
# Determining the RFSI datasets
#training
train_rfsi <- cbind(train, nearest_train)
# validation
test_rfsi <- cbind(test, nearest_test)
# control
names(train_rfsi)

```

## 10. Create the RFSI formula

```{r}
# Creating the RFSI formula 
dist_vars <- paste0("dist", 1:10)
obs_vars <- paste0("obs", 1:10)
rfsi_vars <- c(covars, dist_vars, obs_vars)
rfsi_formula <- reformulate(rfsi_vars, response = "zinc")
rfsi_formula
```

## 11. Fit the RFSO model 

Fit Random Forest using both environmental covariates and spatial neighbour variables.

```{r}
# Learning RFSI model
rfsi_model <- ranger(
  formula = rfsi_formula,
  data = train_rfsi,
  num.trees = 500,
  seed = 42
)
rfsi_pred <- predict(rfsi_model, data = test_rfsi)$predictions

head(data.frame(
    observed = test$zinc,
    RF = rf_pred,
    RFSI = rfsi_pred
))
```
At this stage, predictions from both approaches for Fold 1.

## 12. Run the complete 5-fold cross-validation

```{r}
# 5 fold cycle process
cv_results <- data.frame()
for (fold in 1:5) {
  
  # --------------------
  # Training / validation
  # --------------------
  
  train_id <- which(fold_id != fold)
  test_id  <- which(fold_id == fold)
  
  train <- meuse[train_id, ]
  test  <- meuse[test_id, ]
  
  
  # =========================================================
  # RF
  # =========================================================
  
  rf_model <- ranger(
    zinc ~ dist + ffreq + soil,
    data = train,
    num.trees = 500,
    seed = 42 + fold
  )
  
  rf_pred <- predict(rf_model, data = test)$predictions
  
  
  # =========================================================
  # RFSI
  # =========================================================
  
  nearest_train <- near.obs(
    locations = train,
    locations.x.y = c("x", "y"),
    observations = train,
    observations.x.y = c("x", "y"),
    obs.col = "zinc",
    n.obs = 10,
    rm.dupl = TRUE
  )
  
  nearest_test <- near.obs(
    locations = test,
    locations.x.y = c("x", "y"),
    observations = train,
    observations.x.y = c("x", "y"),
    obs.col = "zinc",
    n.obs = 10,
    rm.dupl = TRUE
  )
  
  train_rfsi <- cbind(train, nearest_train)
  test_rfsi <- cbind(test, nearest_test)
  
  rfsi_model <- ranger(
    formula = rfsi_formula,
    data = train_rfsi,
    num.trees = 500,
    seed = 42 + fold
  )
  
  rfsi_pred <- predict(rfsi_model, data = test_rfsi)$predictions
  
  
  # =========================================================

  # =========================================================
  
  fold_results <- data.frame(
    fold = fold,
    observed = test$zinc,
    RF = rf_pred,
    RFSI = rfsi_pred
  )
  
  cv_results <- rbind(cv_results, fold_results)
}

```
## 13. Calculate RMSE

RMSE measures the average magnitude of prediction errors while giving relatively high weight to large errors.

```{r}
# Calculation the RMSE
rmse <- function(obs, pred) {
  sqrt(mean((obs - pred)^2))
}
# RF
rmse_rf <- rmse(cv_results$observed, cv_results$RF)
# RFSI 
rmse_rfsi <- rmse(cv_results$observed,cv_results$RFSI)

```
## 14. Calculate MAE

MAE represent the average absolute prediction error.

```{r}
# calculating MAE 
mae <- function(obs, pred) {
  mean(abs(obs - pred))
}

mae_rf <- mae(cv_results$observed,cv_results$RF)
mae_rfsi <- mae(cv_results$observed, cv_results$RFSI)

```

## 15. Calculate R2

R² describes how closely predicted and observed values are associated.

```{r}

# Calculating R2
r2 <- function(obs, pred) {
  cor(obs, pred)^2
}

r2_rf <- r2(cv_results$observed, cv_results$RF)
r2_rfsi <- r2(cv_results$observed, cv_results$RFSI)

```

Higher R² indicates better agreement between observed and predicted values.

## 16. Create the performance table

Combine the metrics into one table.

```{r}
# Result figures
performance <- data.frame(
  Model = c("RF", "RFSI"),
  RMSE = c(rmse_rf, rmse_rfsi),
  MAE = c(mae_rf, mae_rfsi),
  R2 = c(r2_rf, r2_rfsi)
)
performance
```
## 17. Calculate the relative improvement of RFSI

The percentage reduction in RMSE can make the comparison easier to interpret.

```{r}
# Calculating the percentage improvement in the RFSI.

rmse_improvement <- 100 * (rmse_rf - rmse_rfsi) / rmse_rf
rmse_improvement
```
If the value is negative, standart RF performed better.

## 18. Plot observed versus predicted values

```{r}
# 18.Observed-predicted plot
plot_data <- cv_results |>
  pivot_longer(
    cols = c(RF, RFSI),
    names_to = "model",
    values_to = "predicted"
  )
ggplot(plot_data, aes(x = observed, y = predicted)) +
  geom_point(alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, linetype = 2) +
  facet_wrap(~ model) +
  coord_equal() +
  theme_minimal() +
  labs(
    x = "Observed zinc (mg/kg)",
    y = "Predicted zinc (mg/kg)",
    title = "5-fold cross-validation"
  )
```
Points closer to the 1:1 line indicate better predictions.

## 19. Compare absolute prediction errors

```{r}
# Comparing the errors
error_data <- cv_results |>
  mutate(
    error_RF = abs(observed - RF),
    error_RFSI = abs(observed - RFSI)
  )
error_long <- error_data |>
  select(error_RF, error_RFSI) |>
  pivot_longer(everything(), names_to = "model", values_to = "absolute_error")

ggplot(error_long, aes(x = model, y = absolute_error)) +
  geom_boxplot() +
  theme_minimal() +
  labs(
    x = NULL,
    y = "Absolute error (mg/kg)",
    title = "RF vs RFSI prediction errors"
  )
```
This figure shows whether RFSI generally produces smaller errors than standart RF. 

## 20. Compare errors statistically

Because both models predict exactly the same validation observations, their errors are paired.

```{r}
# Statistical difference between two models
wilcox.test(error_data$error_RF, error_data$error_RFSI, paired = TRUE)
```
However, this result should be interpreted cautiously because spatial observations are not necessarily statistically independent.
Therefore, the primary evaluation should still rely on:
RMSE
MAE
R²
consistency among the five folds
rather than only on the p-value.

## 21. Calculate performance separately for each fold 

It is useful to check whether one method consistently performs better or whether the overall difference is driven by only one cross-validation fold.
```{r}
# Looking at the performance on a fold basis

fold_metrics <- cv_results |>
group_by(fold) |>
  summarise(
    RMSE_RF = sqrt(mean((observed - RF)^2)),
    RMSE_RFSI = sqrt(mean((observed - RFSI)^2)),
    MAE_RF = mean(abs(observed - RF)),
    MAE_RFSI = mean(abs(observed - RFSI))
  )
fold_metrics
```

## 22. Fit the final Random Forest model

Cross-validation is used to evaluate model performance.

After evaluation, a final model can be trained using all available observations to create the prediction map.

```{r}
# Last model
rf_final <- ranger(
  zinc ~ dist + ffreq + soil,
  data = meuse,
  num.trees = 500,
  seed = 42
)
meuse.grid$zinc_RF <- predict(rf_final, data = meuse.grid)$predictions
```

## 23. Fit the final RFSI model 
First calculate nearest neighbours among the complete set of observations.
```{r}
# Fit the final RFSI model 
nearest_meuse <- near.obs(
  locations = meuse,
  locations.x.y = c("x", "y"),
  observations = meuse,
  observations.x.y = c("x", "y"),
  obs.col = "zinc",
  n.obs = 10,
  rm.dupl = TRUE
)

meuse_rfsi <- cbind(meuse, nearest_meuse)

rfsi_final <- ranger(
  formula = rfsi_formula,
  data = meuse_rfsi,
  num.trees = 500,
  seed = 42
)

```

## 24. Calculate nearest neighbours for the prediction grid

For every grid cell, identify the ten nearest Meuse observations.

```{r}
# Calculate nearest neighbours for the prediction grid
nearest_grid <- near.obs(
  locations = meuse.grid,
  locations.x.y = c("x", "y"),
  observations = meuse,
  observations.x.y = c("x", "y"),
  obs.col = "zinc",
  n.obs = 10,
  rm.dupl = TRUE
)
grid_rfsi <- cbind(meuse.grid, nearest_grid)
meuse.grid$zinc_RFSI <- predict(rfsi_final, data = grid_rfsi)$predictions
```

## 25. Calculate differences between the two prediction maps

The assignment specifically asks for a visual comparison of prediction differences.

```{r}
# Calculate differences between the two prediction maps 
meuse.grid$diff <- meuse.grid$zinc_RFSI - meuse.grid$zinc_RF

```

## 26. Plot the Random Forest prediction map

```{r}
# Plot the Random Forest prediction map
ggplot(meuse.grid, aes(x = x, y = y, fill = zinc_RF)) +
  geom_raster() +
  geom_point(data = meuse, aes(x = x, y = y), inherit.aes = FALSE, size = 0.5) +
  coord_equal() +
  theme_minimal() +
  labs(fill = "Zinc", title = "Random Forest prediction")
```

## 27. Plot the RFSI prediction map

```{r}
# Plot the RFSI prediction map
ggplot(meuse.grid, aes(x = x, y = y, fill = zinc_RFSI)) +
  geom_raster() +
  geom_point(data = meuse, aes(x = x, y = y), inherit.aes = FALSE, size = 0.5) +
  coord_equal() +
  theme_minimal() +
  labs(fill = "Zinc", title = "RFSI prediction")
```

## 28. Plot the differences between RFSI and RF

This is probably one of the most important figures for the assignment.

```{r}
# Plot the difference between RFSI and RF
ggplot(meuse.grid, aes(x = x, y = y, fill = diff)) +
  geom_raster() +
  geom_point(data = meuse, aes(x = x, y = y), inherit.aes = FALSE, size = 0.5) +
  coord_equal() +
  scale_fill_gradient2(midpoint = 0) +
  theme_minimal() +
  labs(fill = "RFSI - RF", title = "Difference between RFSI and RF predictions")
```
This map allows us to identify the locations where explicitly incorporating spatial neighbourhood information changes the predicted zinc concentrations.

## References

Sekulić, A., Kilibarda, M., Heuvelink, G. B. M., Nikolić, M., & Bajat, B. (2020). Random Forest Spatial Interpolation. Remote Sensing, 12(10), 1687. https://doi.org/10.3390/rs12101687

Hengl, T., Parente, L., & Bonannella, C. (2022). Spatial and Spatiotemporal Interpolation / Prediction using Ensemble Machine Learning (Version v0.1). Zenodo. https://doi.org/10.5281/zenodo.5894924
